---
title: "2021 Texas Power Crisis"
output: html_document
date: "2023-12-15"
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Background Information

In February 2021, Texas suffered a major power power crisis with more
than 4.5 million homes and businesses were left without power for
several days[^1]. Historically power outages have disproportionately
affected people of color and low income communities. They take longer to
recover from the disaster because of fewer resources available to them.

[^1]: <https://www.texastribune.org/2021/02/20/texas-power-grid-winterize/>

I am interested in investigating if socioeconomic status played a factor
in community recovery from the power outage.

## Data

My analysis utilized the following data sets:

1.  [Visible Infrared Imaging Radiometer
    Suite](https://ladsweb.modaps.eosdis.nasa.gov/missions-and-measurements/products/VNP46A1/)

Remotely-sensed night light raster I am interested in tiles from
2021-02-07 and 2021-02-16.

2.  [OpenStreetMap](https://www.openstreetmap.org/#map=4/38.01/-95.84):

This contains - Subset of roads that intersect the Houston metropolitan
area - Houses in the Houston Metropolitan area

3.  [U.S. Census Bureau's American Community
    Survey](https://www.census.gov/programs-surveys/acs)

Contains socioeconomic data for each census block

## Data Preparation

#### Import Libraries

Load the necessary libraries. I will mainly be working with `stars`,
`sf` and `ggplot2`.

```{r message=FALSE}
# Libraries
library(tmap)
library(sf)
library(terra)
library(raster)
library(stars)
library(ggplot2)
library(dplyr)
library(ggspatial)
library(patchwork)
```

### Night Lights

First import all of the required data using the `stars` package. Then it
can be conjoined into a single composite.

```{r include=TRUE}
## Import tiles for 2021-02-07 and 2021-02-16
# tile h08v05, collected on 2021-02-07 
feb7_h <- read_stars("~/eds223/assignment-3-hazelvaq/data/VNP46A1/VNP46A1/VNP46A1.A2021038.h08v05.001.2021039064328.tif")

# tile h08v06, collected on 2021-02-07 
feb7_v <-read_stars("~/eds223/assignment-3-hazelvaq/data/VNP46A1/VNP46A1/VNP46A1.A2021038.h08v06.001.2021039064329.tif")

# tile h08v05, collected on 2021-02-16 
feb16_h <- read_stars("~/eds223/assignment-3-hazelvaq/data/VNP46A1/VNP46A1/VNP46A1.A2021047.h08v05.001.2021048091106.tif")

# tile h08v06, collected on 2021-02-16 
feb16_v <- read_stars("~/eds223/assignment-3-hazelvaq/data/VNP46A1/VNP46A1/VNP46A1.A2021047.h08v06.001.2021048091105.tif")

## Convert to stars object
stars_207 <- st_mosaic(feb7_h,feb7_v)
stars_216 <- st_mosaic(feb16_h,feb16_v)

```

I am interested in seeing the difference in night lights intensity from
before and after the storm.

```{r include=TRUE, message=FALSE}
# change in night lights intensity (2/07 - 2/16)
change_lights <- stars_207 - stars_216
```

I assume that any location that experienced a drop of more than 200 NW
cm^-2^sr^-1^ experienced a blackout. Locations that experienced a drop
of *less* than 200 NW cm^-2^sr^-1^ did not experience a blackout and are
assigned `NA` values.

```{r, results='hide'}
# reclassify differences: blackout (> 200 nWcm^-2^sr^-1) and NA (<200) 
dem_rcl <- cut(change_lights,
              breaks=c(-Inf,200,Inf),
              labels = c("blackout","no blackout"))

# assign `NA` to all locations that experienced a drop of *less* than 200 nW cm^-2^sr^-1
change_lights[change_lights <= 200] = NA
```

```{r include=FALSE}
# Vectorize the blackout mask and fix any invalid geometries
blackout_mask <- st_as_sf(change_lights)

blackout_mask <- st_make_valid(blackout_mask)

#View results
#plot(blackout_mask)

# Check if geometries are valid
#st_is_valid(dem_rcl_vector)
```

My area of interest is the Houston metropolitan area and is located
within the following coordinates: (-96.5, 29), (-96.5, 30.5), (-94.5,
30.5), (-94.5, 29). I can crop my blackout mask to this area.

```{r include=TRUE, warning=FALSE}
# Define the Houston metropolitan area 
coord_box <- matrix(c(-96.5,29,-96.5,30.5,-94.5,30.5,-94.5,29,-96.5,29), 
                    ncol = 2, 
                    byrow = TRUE)

# Create a polygon with coordinates
polygon <- st_polygon(list(coord_box))

# Convert the polygon into a simple feature collection
# Assign the CRS to blackout lights tiles
polygon_sf <- st_sfc(polygon, crs = st_crs(blackout_mask))

# Crop (spatially subset) the blackout mask to our region of interest
houston_mask <- st_crop(blackout_mask,polygon_sf)

# Reproject cropped blackout dataset to EPSG:3083
houston_mask <- st_transform(houston_mask, crs = "EPSG:3083")
```

```{r, include=FALSE}
## Plot Houston nigh lights cropped blackout map as a check
tm_shape(houston_mask) +
  tm_polygons(col = "blue") + 
  tm_layout(main.title = "Houston night lights blackout") +
  tm_graticules() +
  tm_compass(size = 0.8, position = c("left","top")) +
  tm_scale_bar(size = 0.5)

```

### Highways 

Highways typically account for most of the night light seen from space.
To remove the potential error of identifying the wrong areas, I will
exclude highways from the blackout mask.

The roads geopackage contains more than highway data so using SQL I
selected just `motorway` data.

```{r include=FALSE, message=FALSE, warning=FALSE, results='hide'}
# SQL query to load only highway data
query <- "SELECT * FROM gis_osm_roads_free_1 WHERE fclass= 'motorway'"

# Load data on highways 
highway <- st_read('data/gis_osm_roads_free_1.gpkg/gis_osm_roads_free_1.gpkg', query = query)

```

Once I have selected the necessary data I can buffer and identify areas
that experience blackouts further than 200m from a highway.

```{r, message=FALSE, warning=FALSE}
# Re-project data to EPSG:3083
highway <- st_transform(highway,crs = "EPSG:3083")

# Identify areas within 200m of all highways
highways_200 <- highway %>% st_buffer(dist = 200) %>%  st_union()

## Identify the areas in houston that experienced blackouts further than 200m from a highway
houston_blackout <- st_difference(houston_mask,highways_200)
```

### Homes Impacted by Blackout

Using SQL I selected only residential buildings from
[OpenStreetMap](https://www.openstreetmap.org/#map=4/38.01/-95.84).
Residential buildings include: residential, apartments, house, static
caravan, and detached.

```{r include=TRUE, message=FALSE, warning=FALSE, results='hide'}
#SQL query to select only residential buildings
query <- "SELECT * FROM gis_osm_buildings_a_free_1 WHERE (type IS NULL AND name IS NULL) OR type in ('residential', 'apartments', 'house', 'static_caravan', 'detached')"

# Import residential buildings
buildings_data <- st_read("~/eds223/assignment-3-hazelvaq/data/gis_osm_buildings_a_free_1.gpkg/gis_osm_buildings_a_free_1.gpkg", query = query)

# Transform buildings data CRS to EPSG:3083
buildings_data <- st_transform(buildings_data, crs = "EPSG:3083")
```

I can now filter to count the number of homes impacted within the
blackout areas.

```{r}
# Filter to homes within blackout areas
homes_blackout <- buildings_data %>% st_filter(houston_blackout,
                          .predicate = st_intersects)

paste("Number of impacted homes:",nrow(homes_blackout))
```

### Socioeconomic Factors

##### load ACS data (10 points)

-   use `st_read()` to load the geodatabase layers\
-   geometries are stored in the `ACS_2019_5YR_TRACT_48_TEXAS` layer\
-   income data is stored in the `X19_INCOME` layer\
-   select the median income field `B19013e1`\
-   hint: reproject data to EPSG:3083\

```{r include=TRUE, message=FALSE, warning=FALSE, results='hide'}
#View ACS layers
st_layers("~/eds223/assignment-3-hazelvaq/data/ACS_2019_5YR_TRACT_48_TEXAS.gdb/ACS_2019_5YR_TRACT_48_TEXAS.gdb")
# Read in geometries and change CRS to EPSG 3083
texas_geometries <- st_read("~/eds223/assignment-3-hazelvaq/data/ACS_2019_5YR_TRACT_48_TEXAS.gdb/ACS_2019_5YR_TRACT_48_TEXAS.gdb", 
                          layer = "ACS_2019_5YR_TRACT_48_TEXAS") %>% 
  st_transform(crs = "EPSG:3083")

# Read in ACS data in X19_INCOME layer and select B19013e1 column
texas_median_income <-
  st_read("~/eds223/assignment-3-hazelvaq/data/ACS_2019_5YR_TRACT_48_TEXAS.gdb/ACS_2019_5YR_TRACT_48_TEXAS.gdb",
    layer = "X19_INCOME") %>% 
  select('GEOID', 'B19013e1') %>% 
  rename("median_income" = 'B19013e1',
         "GEOID_Data" = "GEOID")
```

##### determine which census tracts experienced blackouts (10 points)

-   join the income data to the census tract geometries\
-   hint: make sure to join by geometry ID\
-   spatially join census tract data with buildings determined to be
    impacted by blackouts (-- with our cropped blackout mask and then
    building )\
-   find which census tracts had blackouts\

```{r include=TRUE}
# Join the income data to the census tract geometries 
income_geom <- left_join(texas_geometries,texas_median_income, by = "GEOID_Data") %>% st_transform(crs = "EPSG:3083")

# Check the shape of this variable
#plot(income_geom$Shape) Thus us texas
## Check there are no duplicates in new shape
#unique(duplicated(income_geom$GEOID))

```

I want to crop the recently joined income census tract to Houston
boundaries. I crop the income_geom with the polygon of Houston I made
before.

```{r, warning=FALSE}
# Check if CRS's match
#st_crs(income_geom) == st_crs(polygon_sf)

# Transform Houston polygon CRS to match income_geom 
polygon_sf <- st_transform(polygon_sf, crs = st_crs(income_geom))

# Crop census income data to just Houston area
houston_geom <- st_crop(income_geom,polygon_sf)

# Plot to check shape is now only representative of Houston
#plot(houston_income$Shape)

## spatially join census tract data with buildings determined to be impacted by blackouts
houston_blackout <- st_join(houston_geom,homes_blackout, left = FALSE)

## Census tracts had blackouts - by unique census track
paste("Number of unique census tracks had blackouts:",length(unique(houston_blackout$GEOID)))
```

##### Compare incomes of impacted tracts to unimpacted tracts

-   create a map of median income by census tract, designating which
    tracts had blackouts
-   plot the distribution of income in impacted and unimpacted tracts
-   write approx. 100 words summarizing your results and discussing any
    limitations to this study

```{r, warning=FALSE}
#Create a map of median income by census tract
# Census tracts that had blackouts are designated as red
ggplot() +
  geom_sf(data = houston_geom, aes(fill = median_income)) +
  geom_sf(
    data = houston_blackout,
    aes(color = 'red'),
    fill = NA,
    show.legend = 'abs'
  ) +
  scale_fill_viridis_c() +  
  scale_color_identity(guide = "legend",name = "",labels = "Blackout area", breaks = 'red') +
  annotation_scale(location = 'bl', width = 0.1) +
  annotation_north_arrow(location = 'br',height = unit(.8, "cm"),
  width = unit(.8, "cm"), style = north_arrow_fancy_orienteering()) +
  labs(title = "Houston Median Income Across Census Tracks with Blackouts") +
  theme_minimal() 

```

```{r}
#Plot the distribution of income in impacted and unimpacted tracts
# Simplify and clean Houston census tracks in blackout zone
houston_blackout_dist <- houston_blackout %>% 
  select(GEOID,median_income) %>% unique()

# Income distribution on impacted census tracts
p1 <- ggplot(houston_blackout_dist, aes(x = median_income)) +
  geom_histogram(bins = 20, fill = 'orange', color = 'black') +
  labs(x = 'Median Income', y ='Count', title = 'Income Distribution of Impacted Census Tracts') + ylim(0,150) + theme(text=element_text(size=7))


```

```{r, warning=FALSE}
# Income distribution of unimpacted tracts
#Clean up Houston geom 
houston_geom_dist <- houston_geom %>% select(GEOID,median_income) %>% unique

# Identify the census track that were not affected by the blackout
houston_non <- setdiff(houston_geom_dist,houston_blackout_dist)

## Check difference is correct
#nrow(houston_geom_dist) == nrow(houston_non) + nrow(houston_blackout_dist)

# Plot income distribution of census tracts not affected by the blackout
p2 <- ggplot(data = houston_non, aes(x = median_income)) +
  geom_histogram(bins = 20, fill = '#4398F9', color = 'black')+
  ylim(0,150) + labs(x = "Median Income",y = "Count", title = "Income Distribution of Unimpacted Census Tracts") + theme(text=element_text(size=7))


p1+p2
```

### Conclusion

The income distribution of the impacted census tract has a similar
distribution of the un-impacted census tract. They both display a right
skew with the majority of the distribution being in the lower income
range. During this storm income level of the census tract did not
display a large effect based solely on the graphs generated. There are
some limitations to this study such as the limitation of spatial data in
the days during the storm because of the cloud coverage. The raster data
I analyzed is during the second storm and does not account for the third
storm. Also, there is the issue of the year the ACS data represents. The
income demographic are from 2019, these could have changed since then
and potentially do not accurately represent the income distribution of
Houston during the storm.
